Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems

### 2024-06-28 2024/OPTIMA/OPTIMA_2024_MCO.tex
minor changes

After the construction of the separating hyperplane, for each trial point the $d_i$ distance is calculated and normalized according to the following expression:
\begin{equation}
    \label{eq:17}
d'_i=
\begin{cases}
  \begin{matrix}
     d_i / d_{max}, & d_i > 0 \\
     -d_i / d_{min}, & d_i < 0 
  \end{matrix}
\end{cases}, 
0 \leq i \leq k-1,
\end{equation}
where $d_i$ is the distance from the $i$-th point to the hyperplane, $d_{max}$, $d_{min}$ are maximal and minimal distances to the hyperplane.

@@

For creating the penalty term $R_{ps}(i)$ from (\ref{eq:16}) that corresponds better to the estimation of the Pareto set, the following new ideas are suggested.

\begin{enumerate}
	\item Probabilistic approach with application of machine learning models.   
	\item Automatic adjustment of class weights when building a machine learning model.
	\item Transformation of machine learning probabilities.
\end{enumerate}

@@

The color background displays the behavior of the penalty $R_{ps}(i)$ as a function of the criteria $f1$ and $f2$. Fig. \ref{fig:1}(a) shows the function $R_{ps}(i)$ for the method ML\_MGSA\_Dist. Fig. \ref{fig:1}(b), \ref{fig:1}(c), and \ref{fig:1}(d) demonstrate $R_{ps}(i)$ for the different cases of the probabilistic models. Fig. \ref{fig:1}(b) corresponds to the model with linear kernel (ML\_MGSA\_Prob\_Linear), Fig. \ref{fig:1}(c) reflects results for the model with polynomial kernel (ML\_MGSA\_Prob\_Poly), and Fig. \ref{fig:1}(d) is related to RBF-based kernel (ML\_MGSA\_Prob\_RBF).

The number of trials carried out in each experiment is indicated in parentheses after the name of the method.

@@

Computational experiments were conducted for each method and each model in order to assess the effectiveness of using these techniques in the global search algorithm. Fig. \ref{fig:2} demonstrates the visualization of $R_{ps}(i)$ similar to Fig. \ref{fig:1}.

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{fig2.png}
\caption{Heatmap of the function $R_{ps}(i)$ in the criteria space for algorithms with different techniques of building models after transformation (\ref{eq:18}).} 
\label{fig:2}
\end{figure}

It is easy to notice that new penalty functions (see Fig. \ref{fig:2}. (b)-(d)) describe better the Pareto set than functions built on the base of separating hyperplane. A detailed comparison of approaches to different computation of the characteristic (\ref{eq:16}) is given in Section \ref{sec:5}.

>>>>

After the construction of the separating hyperplane, for each trial point the $d_i$ distance is calculated and normalized according to the following expression:
\begin{equation}
    \label{eq:17}
d'_i=
\begin{cases}
  \begin{matrix}
     d_i / d_{max}, & d_i > 0 \\
     -d_i / d_{min}, & d_i < 0 
  \end{matrix}
\end{cases}, 
0 \leq i \leq k-1,
\end{equation}
where $d_i$ is the distance from the $i$-th point to the hyperplane, $d_{max}$, $d_{min}$ are maximum and minimum distances to the hyperplane.

@@

For creating the penalty term $R_{ps}(i)$ from (\ref{eq:16}) that corresponds better to the estimation of the Pareto set, the following new ideas are suggested: 
\begin{enumerate}
	\item Probabilistic approach with application of machine learning models;   
	\item Automatic adjustment of class weights when building a machine learning model;
	\item Transformation of machine learning probabilities.
\end{enumerate}

@@

The color background displays the behavior of the penalty $R_{ps}(i)$ as a function of the criteria $f_1(y)$ and $f_2(y)$. Fig. \ref{fig:1}(a) shows the function $R_{ps}(i)$ for the method ML\_MGSA\_Dist. Fig. \ref{fig:1}(b,c,d) demonstrate $R_{ps}(i)$ for the different cases of the probabilistic models. Fig. \ref{fig:1}(b) corresponds to the model with linear kernel (ML\_MGSA\_Prob\_Linear), Fig. \ref{fig:1}(c) reflects results for the model with polynomial kernel (ML\_MGSA\_Prob\_Poly), and Fig. \ref{fig:1}(d) is related to RBF-based kernel (ML\_MGSA\_Prob\_RBF).
The number of trials carried out in each experiment is indicated in parentheses after the name of the method.

@@

Computational experiments were conducted for each method and each model in order to assess the effectiveness of using these techniques in the global search algorithm. Fig. \ref{fig:2} demonstrates the visualization of $R_{ps}(i)$ similar to Fig. \ref{fig:1}.
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{fig2.png}
\caption{Heatmap of the function $R_{ps}(i)$ in the criteria space for algorithms with different techniques of building models after transformation (\ref{eq:18}).} 
\label{fig:2}
\end{figure}

Note that new penalty functions (see Fig. \ref{fig:2}(b,c,d)) describe better the Pareto set than functions built on the base of separating hyperplane. A detailed comparison of approaches to different computation of the characteristic (\ref{eq:16}) is given in Section \ref{sec:5}.
